% !TEX root = ../thesis.tex
%
\pdfbookmark[0]{Acknowledgement}{Acknowledgement}
\chapter*{Acknowledgement}
\label{sec:acknowledgement}
\vspace*{-10mm}

I am thankful that I could be a member of the group on \emph{Programming Languages and Compiler Construction} led by my advisor Michael Hanus for the last five and a half years.

My most profound thanks deserves Jan for having the greatest impact in my growth as a researcher and as a person.
We can look back at 5 years of intensive research corporations including all the highs and lows our profession comes with: paper rejections and acceptances remain stable, some ideas worked out and some did not.
In the end, a problem shared is a problem halved, and an achievement shared is a delight doubled.
During my first years as PhD student I have a lot of great memories working with Jan and Nikita.
The two years I shared my working environment with Nikita, he was my main contact for all occasions at work: lunch, Haskell, teaching, and research in general.

As usually, when great colleagues leave, there are new people coming into your life.
In this respect, I thank my colleagues Finn and Niels for keeping me sane, especially in the final year of my PhD.
It is with first and foremost joy and appreciation that I look back at our joint work on papers and discussions about research topics in general.
I also loved hanging out with you at our lunch and occasional coffee breaks, which we shared with Johannes, Christian and Frank.
Moreover, I am grateful for the pleasant working environment on the \emph{7th floor} in general, including the people not mentioned so far: JR, Ina, Renate, Rudolf, and Ulrike as well as the Archaeoinformatics group.

My special thanks to Julia, Jan (\emph{Herr Bracker}), Ole, Marcellus, Kai, Nikita, Niels, and Finn for reading drafts of this thesis and providing me with helpful feedback to improve my writing.
And I also like to thank Insa for giving me insights of her work using the proof assistant Coq that led to my own research interest in that area.
Additionally, I was glad to have Janis Voigtl\"ander and Thomas Wilke on my examination committee and am, especially, thankful for Janis' expert comments on the topics of this thesis.

I hope it is without any doubt that I enjoyed teaching a lot of students the ideas of functional programming and I am especially proud of all the students, whose thesis I advised: you all did a great job and gave me the perfect opportunity to deepen my understanding of my own research.

Last but not least, the aforementioned coffee breaks would not have been the same without the great team at the \emph{Caf\'e Lounge}: it was with great pleasure that I could order my chai latte with oat milk just saying ``The usual, please''.